### 2024-04-11 docs/SOSP 2024/intro.tex
intro joint pass

Decision forest models are composed of a large collection of decision trees (100-1000), and inference involves 
traversing down each tree in the forest and aggregating the predictions. Inference is typically done in a batched
setting, where multiple inputs are processed simultaneously.
Despite the simplicity of the model and the availability of multiple sources of coarse grain parallelism (parallelism 
across inputs in a batch and parallelism across trees), existing systems do not consistently scale well across different
models even on the limited set of targets they support. 
This is because existing systems exploit a limited set of optimizations and often specialize the implementation 
to a specific hardware platform, and this limits their portability.

%\TODO{cover Rapids, Tahoe, XGBoost and Treebeard}.  
Evaluation on a diverse set of models highlights that the best implementation often requires a careful combination 
of many optimization strategies. For example, simultaneously traversing multiple trees with many threads, while managing the working set is 
critical to get scalable performance. This requires a combination of techniques like data layout optimizations, 
loop transformations, and memory access optimizations. Existing systems each use specific data layouts for the underlying
tree and use fixed loop structures. 
Parallelism, loop structure and model representation together determine the accesses patterns and the working set. A fixed configuration 
as used by existing systems does not perform consistently well across different models and hardware platforms.
XGBoost~\cite{XGBoost} uses a sparse representation for the model and a loop structure that processes one tree for a block of rows before moving to 
the next tree. RAPIDS FIL~\cite{FIL} uses a reorg representation and partitions trees across a fixed number of threads. Tahoe~\cite{Tahoe} uses a
variation of the reorg representation and has four predefined inference strategies from which it picks one based on an analytical model.
These systems are all library based and are hard to extend. 
\TreebeardOLD{}, the state-of-the-art model compiler for CPUs, supports two fixed loop structures and does not scale well with increasing number of threads.

This paper presents \Treebeard{}, a novel schedule guided compilation infrastructure for decision tree inference on multiple target hardware. \Treebeard{} is able to generate high-performance code for decision tree inference by exploring a large optimization space. The code generation is guided by a \emph{schedule} object that allows us to effectively represent this solution space abstractly. The schedule object is written in \Treebeard{} custom scheduling language that is expressive enough to represent a wide range of implementation strategies. We demonstrate that the language is sufficient to express the various optimizations proposed by prior work, and further generalizes them and incorporates several new optimizations. We also design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.
\TODO{Performance evaluation summary}
\TODO{Oganization}
\subsection{Contributions}
\begin{itemize}
  \item We present the design for a multi-target decision tree compiler infrastructure and implement several optimizations within this framework.
   We are also the first to implement an optimizing compiler for decision tree inference on GPUs.
  \item We identify that an extensive optimization space exists for the problem of decision tree inference. We design a scheduling language that 
  allows us to effectively represent this solution space abstractly. This scheduling language is expressive enough to represent a wide range of  
  implementation strategies proposed by prior work.
  \item To the best of our knowledge, we perform the first extensive characterization of the optimization space for decision tree inference 
  on GPUs. Using some of the characteristics we identify, we design and implement a heuristic that is able to quickly find high-performance 
  schedules for the model being compiled. 
  \item We design and implement a general framework for expressing and optimizing reductions within MLIR. To the best of our knowledge, this is the first
  such framework.
  \item We evaluate our implementation by comparing it against RAPIDs and Tahoe, the state-of-the-art decision tree inference frameworks for GPU and 
  report significant speedups. We also show that our compiler can effectively target different GPUs, including both NVIDIA and AMD GPUs.
\end{itemize}

>>>>

Decision forest models are composed of a large collection of decision trees (100-1000), and inference involves 
traversing down each tree in the forest and aggregating the predictions. Inference is typically done in a batched
setting, where multiple inputs are processed simultaneously.
Despite the simplicity of the model and the availability of multiple sources of coarse grain parallelism (parallelism 
across inputs in a batch and parallelism across trees), existing systems do not consistently scale well across different
models even on the limited set of targets they support. 
%This is because existing systems exploit a limited set of optimizations and often specialize the implementation to a specific hardware platform, and this limits their portability.

%\TODO{cover Rapids, Tahoe, XGBoost and Treebeard}.  
Evaluation on a diverse set of models highlights that the best implementation often requires a careful composition 
of many optimization strategies like data layout optimizations, 
loop transformations, parallelization, and memory access optimizations. 
%For example, simultaneously traversing multiple trees with many threads, while managing the working set is 
%critical to get scalable performance. 
%This requires a combination of techniques like data layout optimizations, loop transformations, and memory access optimizations. 
%Existing systems each use specific data layouts for the underlying tree and use fixed loop structures. 
%Parallelism, loop structure and model representation together determine the accesses patterns and the working set. 
Existing systems today are mostly library based, and only support a predefined combination of optimizations. 
XGBoost~\cite{XGBoost} uses a sparse representation for the model and a loop structure that processes one tree for a block of rows before moving to 
the next tree. RAPIDS FIL~\cite{FIL} uses a reorg representation and partitions trees across a fixed number of threads. Tahoe~\cite{Tahoe} uses a
variation of the reorg representation and has four predefined inference strategies from which it picks one based on an analytical model.
%These systems are all library based and are hard to extend. 
\TreebeardOLD{}, the state-of-the-art model compiler for CPUs, supports two fixed loop structures and does not scale well with increasing number of threads. Additionally, it lacks GPU specific optimizations that are critical to scale performance to massive number of threads.

This paper presents \Treebeard{}, a novel schedule guided compilation infrastructure for decision tree inference on multiple target hardware. \Treebeard{} is able to generate high-performance code for decision tree inference by exploring a large optimization space. The code generation is guided by a \emph{schedule} written in a custom scheduling language that can represent a wide range of implementation strategies. We demonstrate that the language is sufficient to express the various optimizations proposed by prior work, and further generalizes them and incorporates several new optimizations. We also design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.
\TODO{Performance evaluation summary}
\TODO{Oganization}
\subsection{Contributions}
\begin{itemize}
  \item We present the design for a multi-target decision tree compiler infrastructure and implement several optimizations within this framework.
   We are also the first to implement an optimizing compiler for decision tree inference on GPUs.
  \item We identify that an extensive optimization space exists for the problem of decision tree inference. We design a scheduling language that 
  allows us to effectively represent this solution space abstractly. This scheduling language is expressive enough to represent a wide range of  
  implementation strategies proposed by prior work.
  \item To the best of our knowledge, we perform the first extensive characterization of the optimization space for decision tree inference 
  on GPUs. Using some of the characteristics we identify, we design and implement a heuristic that is able to quickly find high-performance 
  schedules for the model being compiled. 
  \item We design and implement a general framework for expressing and optimizing reductions within MLIR. To the best of our knowledge, this is the first
  such framework.
  \item We evaluate our implementation by comparing it against RAPIDs and Tahoe, the state-of-the-art decision tree inference frameworks for GPU and 
  report significant speedups. We also show that our compiler can effectively target different GPUs, including both NVIDIA and AMD GPUs.
\end{itemize}
